sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-08-03 doc/sdmTMB-paper.Rmd
Rename/format

---
title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
preprint: false
author:
  - name: Sean C. Anderson
    affiliation: 1
    corresponding: true
    email: sean.anderson@dfo-mpo.gc.ca
  - name: Eric J. Ward
    affiliation: 2
  - name: Philina A. English
    affiliation: 1
  - name: Lewis A. K. Barnett
    affiliation: 3
affiliation:
  - code: 1
    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - code: 2
    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
  - code: 3
    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
abstract: >
  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to. Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. We anticipate its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
output:
  bookdown::pdf_book:
    template: template-peerj.tex
    keep_tex: true
    pandoc_args: "--highlight=haddock"
bibliography: refs.bib
link-citations: yes
csl: peerj.csl
header-includes:
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
---

<!-- Point 1: set the context for and purpose of the work; -->
<!-- Point 2: indicate the approach and methods; -->
<!-- Point 3: outline the main results; -->
<!-- Point 4: identify the conclusions and the wider implications. -->

<!-- TODO: emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference-->

@@

<!-- File naming format: 'Supplemental [Item] S[number]' e.g. 'Supplemental Data S1'. -->
<!-- Use the following style when citing Supplemental files in the text: -->
<!-- Fig. S1, Table S1, Data S1, Video S1, Article S1, Audio S1. -->

<!-- Introduction -->
<!-- Materials & Methods -->
<!-- Results -->
<!-- Discussion -->
<!-- Conclusions -->
<!-- Acknowledgments -->

# Introduction {-}

@@

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram), Article S1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

<!-- Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time, -->
<!-- the general form of a GLMM is -->
<!-- $g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$, -->
<!-- with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$. -->
<!-- We can partition the random effects $\bm{Z \psi}$ into subcomponents, each of which can optionally be included in the model (Appendix TODO). -->
<!-- Although, in practice it is rare to include all of these components, the full sdmTMB model is -->
<!-- $$ -->
<!-- g(\mu_{\bm{s},t}) = -->
<!-- \bm{X}_{1,\bm{s},t} \bm{\beta} + -->
<!-- \bm{\alpha}_g + -->
<!-- \bm{X}_{2,\bm{s},t} \gamma_t + -->
<!-- \bm{X}_{3,t} \zeta_{\bm{s}} + -->
<!-- \omega_{\bm{s}} + -->
<!-- \epsilon_{\bm{s},t}, -->
<!-- $$ -->
<!-- where -->

<!-- * $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices; -->
<!-- * $\bm{\beta}$ represents a vector of fixed-effect coefficients; -->
<!-- * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; -->
<!-- * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$; -->
<!-- * $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$; -->
<!-- * $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and -->
<!-- * $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field). For the IID case, $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$ (Appendix TODO). -->

@@

(ref:diagram-fig) **Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.**
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Article S1 for a full description of the model.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Article S1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Article S1).
The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

@@

# Introductory example using fish survey data {-}

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
<!-- TODO: Since this is not particular to your package, maybe the e.g. for how to do this should be moved to the appendix? I do explain the transformation for the owl data and added a sentence to the pcod appendix, but maybe still important to include here as well? -->
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

@@

@@

(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
\noindent
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

@@

We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Article S2).

# Example of a spatially varying coefficient with citizen science data  {-}
<!-- # Estimate a spatially varying effect from citizen science data -->

@@

We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Article S3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.

(ref:owl-nao-fig) **Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.**
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NAO unit.

@@

There are several R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with various limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Article S4) and VAST (Article S5).

(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
Lines represent means and ribbons 95% quantiles across 50 random iterations.
Note the log x and y axes.
<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
VAST is similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
Note that spaMM only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Article S6).
In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
mgcv was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA will find inlabru approachable.
Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Article S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages sdmTMB and VAST were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

Speed-wise, sdmTMB (and by association VAST) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than INLA/inlabru across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.
<!-- Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores. -->

<!-- However, with the SPDE, mgcv is typically slower than sdmTMB for similar models without optimized matrix libraries and lacks certain  -->

<!--
There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2019]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
From a fisheries perspective, sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications or support delta/hurdle models.

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig. \@ref(fig:timing), @osgood-zimmerman2021].
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

>>>>

---
title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
author: "|
  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^3$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^4$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    bookdown::word_document2:
    # bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
      keep_tex: true
bibliography: refs.bib
link-citations: yes
csl: ecography.csl
header-includes:
  \usepackage[document]{ragged2e}
  \usepackage{setspace}\doublespacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \setlength{\parskip}{1.0\baselineskip}
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{lscape}
  \widowpenalty10000
  \clubpenalty10000
---

@@

\clearpage

\linenumbers

# sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

# Abstract

Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. 
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. 
Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. 
Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. 
We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). 
Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. 
Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. 
We anticipate that its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.

Keywords:
Gaussian Markov random fields (GMRF),
generalized linear mixed effects models (GLMM),
INLA,
SPDE,
species distribution modelling,
R package,
spatio-temporal or spatial-temporal,
Template Model Builder

\clearpage

# Background {-}

@@

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram), Appendix S1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix S1 for a full description of the model.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix S1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix S1).
The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

@@

# Example using fish survey data {-}

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

\onehalfspacing

@@

\doublespacing

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).

@@

\doublespacing

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
\noindent
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.

\onehalfspacing

@@

\doublespacing

We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Appendix S2).

# Example of a spatially varying coefficient with citizen science data  {-}

@@

\doublespacing

We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix S3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NAO unit.

@@

There are several R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with various limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix S4) and VAST (Appendix S5).

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons 95% quantiles across 50 random iterations.
Note the log x and y axes.
<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
VAST is similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
Note that spaMM only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix S6).
In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
mgcv was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA will find inlabru approachable.
Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages sdmTMB and VAST were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

Speed-wise, sdmTMB (and by association VAST) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than INLA/inlabru across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.

Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
The included TMB `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
